Legacy Data Confounds Genomics Studies

### 2018-10-02 legacy.tex
made recommended changes
added text about hap map and gnomAD sites.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\title{Legacy Data Confounds Modern Genomics Studies}
\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}

@@

While certain mutagens have very distinct mutational signatures, others are more subtle and require much larger datasets to identify because they tend to occur less frequently and with less genomic specificity \citep{Pleasance2010}. \sgcomment{Missing logical link}
Differences in mutational signatures across human populations have recently been reported.
In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}. 
Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.

@@

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2015a} \sgcomment{That would be 2017, no?}.
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, and in essence we would not expect any homozygous alternates for low frequency variants. \sgcomment{this sentence needs work}

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a \textit{reverse} genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
Here we use logistic regression of the genotypes on a quantitative measure of quality (i.e., the genotype is the dependent variable).
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could be as a result of \sgcomment{could result from?} fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.

@@

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels), as shown on Figure \ref{Manhattan} and Figure \ref{indel_Manhattan} \sgcomment{figure missing for me}.

@@

\sgcomment{In order to -> To, in general} In order to investigate the proportion of variants that we have identified as being associated to $Q$ that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
We found that all of the \sgcomment{"all of the" -> "all the" } SNPs we identified as being associated with $Q$ were imputed back in the samples. 
This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual

Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait. 
The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had found variants to have either reached or nearly reached genome wide significance of $ p < 10^{-8}$. \sgcomment{This is misleading! Neither has reached $10^-8$}

@@

We were not able to pinpoint the precise technical source of the discrepancy. Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced with an entirely different platform. \sgcomment{Check that is sanger, if so, say so}. We find that \sgcomment{$99\%$--check} of the suspicious snps are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.

The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 1\% of sites identified in 1kGP present in HapMap.
Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
\caption{Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.  }
\label{PC1_Correlation}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{MostSig}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{TwinsSNP}
\end{figure}

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
\label{Singles}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Genomics Studies}
%\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}

@@

In addition to somatic mutational signatures, there has been recent interest in population variation in germline mutational signatures, which can be revealed in large sequencing panels. 
In 2015, Harris reported 50\% more TCC ${\rightarrow}$ TTC mutations in European populations compared to African populations, and this was replicated in a different cohort in 2017 \citep{Harris2015a, Harris2017a, Mathieson2017a}. 
Strong population enrichments of a mutational signature suggests important genetic or environmental differences in the history of each population \citep{Harris2015a, Harris2017a}. 
Harris and Pritchard further identified distinct mutational spectra across a range of populations, which were further examined in a recent publication by Aikens et al. \citep{Harris2017a,Aikens2018}.

@@

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP \citep{Harris2017a}.
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect.

These mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
Whereas Hardy-Weinberg testing is efficient at detecting high frequency spurious variants, it has little power to filter out rare variants which, in aggregate, can contribute to mutational signatures.

@@

To identify SNPs that are likely to reproduce poorly across cohorts (without having access to a second cohort), we performed a reverse genome wide association study (GWAS) in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use linear regression of a phenotype on a genotype. 
Here we use logistic regression of the genotypes on a quantitative measure of quality (i.e., the genotype is the dependent variable).
This approach identifies 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$.
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations.
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

	\subsection{Identifying Suspicious SNPs in the 1kGP}
The results of the reverse GWAS performed on the Japanese from the 1kGP lead us to consider whether any other populations in the 1kGP also had similar issues with data quality.
The $Q$ of individuals over time shows that the sequencing done in the early phases of the 1kGP was more variable and overall tended to include lower quality sequencing data (Figure \ref{MapQual}).
This variability could result from fluctuating sequence platform protocols or variation between sequencing centres.
There were a number of protocols used to prepare the samples for sequencing, as well as a number of sequencing technologies used over the course of the data production.

@@

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels) (Figure \ref{Manhattan} and Figure \ref{indels_Manhattan}).

@@

In general, to investigate the proportion of suspicious SNPs that are being imputed, we submitted genotype data from chromosome 1 and 2 of 1kGP individuals to the Michigan Imputation Server.
We found that all the SNPs we identified as being associated with $Q$ were imputed back in the samples. 
This suggests that the imputation reference panel likely includes individuals with low $Q$, and the dubious variants will be imputed in individuals who most closely match the low-quality individual

Once we identified SNPs that were clearly associated with low $Q$, we searched the literature for any GWAS that might have reported these dubious variants as being significantly associated with some biological trait. 
The NHGRI-EBI Catalog of published genome-wide association studies identified two recent publications that had described variants that \todo{nearly reached genome wide significance} of $ p < 10^{-8}$.

@@

We were not able to pinpoint the precise technical source of the discrepancy. 
Sequencer-specific errors would be expected to be absent from Hapmap, which was sequenced using a combination of array based technologies and PCR targeted sequencing \citep{HapMap2005}. 
We find that $99\%$ of the suspicious SNPs are indeed absent from HapMap, which is consistent with the expected false positive rate in our test.

The presence of some of these SNPs in HapMap suggests a cell line effect; although, this result is not very convincing with only 42 variants of sites identified in 1kGP present in HapMap (Figure \ref{HapMap_GnomAD}).
We also note that 164 of the suspicious variants are present in the GnomAD database (Figure \ref{HapMap_GnomAD}).
Indeed, further inquiries into the details of sample preparation and data processing would be required to identify the source of this signal.

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/AllDeviances.jpg}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/PC1_Correlation.jpg}
\caption{Regressions against the average quality per mapped bases $Q$ per individual is the best correlate with prevalence of the  *AC${\rightarrow}$*CC mutational signature in 1kGP, with individuals with low-quality data showing elevated rates of the signature.  }
\label{PC1_Correlation}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_mostSig2.jpg}
\caption{Logistic regression of the most significantly associated SNP rs75254682 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{MostSig}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot.jpg}
\caption{Logistic regression of rs6057648 for the populations with significant association to the average quality per mapped bases $Q$.}
\label{TwinsSNP}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
\label{Singles}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/Hap_GnomAD.jpg}
\caption{Site frequency spectrum of suspicious SNPs found in HapMap and GnomAD. The blue dot in the HapMap plot is the joint frequency of a variant from a published GWAS. There are 42 suspicious SNPs present in the HapMap database and 162 suspicious SNPs present in the GnomAD database.}
\label{HapMap_GnomAD}
\end{figure}
